Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-01-16 src/docs_paper/paper.Rmd
Rerender paper and appendix

Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution to make targetted interventions.
Producing these estimates is a challenging task, as all available data sources have shortcomings which must be overcome.
Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they only occur infrequently.
Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or biased samples of the population.
To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
Concurrent modelling of data mitigates the limitations of any single data source, increases statistical power and prompts investigation into any conflicting information.
Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and generate estimates in a yearly process supported by UNAIDS.

The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
For the Naomi model, we use this option to integrate out the latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
As such, we have been motivated to find a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
The central innovation of @rue2009approximate is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
`aghq` is designed to naturally interface with `TMB`, so use of our method is simple when provided a \textsf{C++} user template for the log-posterior.

The remainder of this paper is organised as follows.
In Section \ref{sec:naomi} we describe a simplified version of the Naomi model that we consider in this paper.
Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
As a case-study, we fit the simplified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
In this section, we also demonstrate a Bayesian workflow, illustrating that these concepts are not only applicable in a simulation-based inference setting.
Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how our method might be used in other models, and directions for future research.

@@

* We developed an approximate Bayesian inference algorithm to solve a challenging problem in the small-area estimation of HIV in low resource settings
* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible
* We demonstrated a Bayesian workflow for deterministic inference methods

>>>>

Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution to make targetted interventions.
Producing these estimates is a challenging task beccause all available data sources have shortcomings which must be overcome.
Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they occur only infrequently.
Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or biased samples of the population.
To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
Drawing on expertise closest to the data generating process, improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.

The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
This has motivated us to look for a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
The central innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
Since `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.

The remainder of this paper is organised as follows.
In Section \ref{sec:naomi} we describe a simplified version of the Naomi model that we consider in this paper.
Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
As a case-study, we fit the simplified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how we anticipate our method might be useful for other models, and directions for future research.

@@

* We developed an approximate Bayesian inference algorithm to solve a challenging problem in the small-area estimation of HIV in low resource settings
* Following further testing, we anticipate adding our inference method as an option for 
* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible. We are excited about exploration of its accuracy for the following challenging model structures which could not previously be fit: list of model types here.
* We demonstrated a Bayesian workflow for deterministic inference methods
